Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-06-18 writing/DraftExposurePaper/DraftExposurePaper.Rmd
Work on paper draft

```{r echo = FALSE}
library(hurricaneexposure)
library(knitr)
library(gridExtra)
```

@@

Many outcomes of interest are available at county-level aggregations (e.g., counts of health outcomes: @Grabich2016: premature births and low birth weights; @Grabich2015: birth rates). Further, decisions and policies to prepare for and respond to storms are often undertaken at the county level [@Zandbergen2009]. Some studies suggest methods to determine exposure, including inland exposure, to tropical storms at the county-level over many years [@Zandbergen2009]. Here, we offer tools, through the open-source `hurricaneexposure` package, to determine county-level exposure to tropical storms using several different metrics (distance, rainfall, wind). In this paper, we describe how this software package assesses exposure to tropical storms at the county level, provide an analysis of the advantages and disadvantages of different strategies for assessing county-level exposure, and explore variation in exposure assessments based on using different exposure metrics.

Current hurricane exposure assignments are often nonspecific [GBA-- I found this wording a bit unclear] and may lead to missclassification of exposure at the county-level [@Grabich2016]. Researchers have found heterogeneity in exposure assignments of counties, both within and between storms, when different metrics of exposure are used [@Grabich2016; @Grabich2015]. Such missclassification can have important implications of assessments of the public health and economic impacts of tropical storms.

@@

If exposure to tropical storms over multiple storms and years can be assessed, these exposure datasets can be joined with other time series to explore the impacts of tropical storms. For example, daily counts of human health outcomes in environmental epidemiology studies are often available aggregated at the county level, and such data has often been paired with time series of environmental exposures (e.g., air pollution, temperature) to determine associated risks.

@@

We collected "best tracks" data on hurricane tracks for Atlantic basin storms between 1988 and 2014 from the extended best tracks database. This dataset is based on a poststorm assessment of each storm conducted by the United States National Hurricane Center (NHC) and incorporates data from a variety of sources, including satellite data and, when available, aircraft reconnaisance data [@Landsea2013]. This data gives time stamps for each observation in Coordinated Universal Time (UTC; also known as Zulu Time, sometimes indicated by "Z").

@@

We calculated the distance between each county's center and each of the 15-minute-interval estimates of a storm's location. To do this, we used the Great Circle method to calculate the distance between pairs of latitude and longitude coordinates, using the R package `sp`. For each county, we used the population mean center, based on the U.S. Census's 2010 Decennial Census, as the center location of the county. From these distance calculations, we identified the date-time and distance between the storm track and the county center at the 15-minute interval when the storm was closest to the county center. We used the `countytimezones` R package to convert each date-time of the closest storm approach for a county to the county's local time zone and identified the local date when the storm was closest to the county. This "closest date" was used to pair the distance estimates with rainfall estimates.

@@

NLDAS does not provide information on precipitation over the ocean, preventing its use in assessing the develpment of the storm prior to landfall [@Villarini2011]. This point is moot for this analysis.

@@

```{r echo = FALSE, fig.width = 6, fig.height = 8}
a <- map_counties(storm = "Floyd-1999", metric = "rainfall")
a <- map_tracks(storms = "Floyd-1999", plot_object = a, plot_points = FALSE)
b <- map_counties(storm = "Floyd-1999", metric = "distance")
b <- map_tracks(storms = "Floyd-1999", plot_object = b, plot_points = FALSE)
grid.arrange(a, b, ncol = 1)
```

@@

There are a number of limitations to using distance to assess exposure to tropical storms. Tropical storms vary in size and intensity, and a measure of distance from the storm track will not incorporate these differences and so could misclassify exposure both in terms of generating false positives (counties close to the storm track of very mild or small-radius storms) and false negatives (e.g., during very large or intense storms). In particular, the forces of a storm tend to be distributed around the center in a non-symmetrical way [citation], while distance-based exposure metrics that use buffers tend to use an equal buffer distance on each side of the storm track (e.g., [citations]).

@@

@@

Exposure classification based on rainfall has some advantages. For example, it allows the identification of exposed counties that are inland, rather than coastal, but that were affected by heavy rainfall during the storm. Often, storm-related deaths are associated with inland flooding. In fact, over half of the hurricane-related fatalities from 1970 to 1999 were a result of freshwater flooding, accounting for the vast majority of deaths in inland counties [@Rappaport2000]. In another analysis, researchers found that 79\% of freshwater-drowning fatalities occured in inland counties,[@Czajkowski2011] further stressing the importance of rainfall in hurricane risk analysis. Storms can produce a lot of rain especially in certain topographies, like near mountains, so counties that are well inland can sometimes experience more extreme rain that other counties at similar distances from the storm's track.

@@

@@

>>>>

```{r echo = FALSE}
library(hurricaneexposure)
library(knitr)
library(ggplot2)
library(gridExtra)
```

@@

Many outcomes of interest are available at county-level aggregations (e.g., counts of health outcomes: @Grabich2016: premature births and low birth weights; @Grabich2015: birth rates; @Rappaport2000: direct hurricane-related deaths). Further, decisions and policies to prepare for and respond to storms are often undertaken at the county level [@Zandbergen2009; @Rappaport2000]. Some studies suggest methods to determine exposure, including inland exposure, to tropical storms at the county-level over many years [@Zandbergen2009]. Here, we offer tools, through the open-source `hurricaneexposure` package, to determine county-level exposure to tropical storms using several different metrics (distance, rainfall, wind). In this paper, we describe how this software package assesses exposure to tropical storms at the county level, provide an analysis of the advantages and disadvantages of different strategies for assessing county-level exposure, and explore variation in exposure assessments based on using different exposure metrics.

Current hurricane exposure assignments are often nonspecific [GBA-- I found this wording a bit unclear] and may lead to missclassification of exposure at the county-level [@Grabich2016]. Researchers have found heterogeneity in exposure assignments of counties, both within and between storms, when different metrics of exposure are used [@Grabich2016; @Grabich2015]. Such missclassification can have important consequences in assessments of the public health and economic impacts of tropical storms.

@@

If exposure to tropical storms over multiple storms and years can be assessed, these exposure datasets can be joined with other time series to explore the impacts of tropical storms. For example, daily counts of human health outcomes in environmental epidemiology studies are often available aggregated at the county level, and such data has often been paired with time series of environmental exposures (e.g., air pollution, temperature) to determine associated risks. Databases of storm impacts, including direct deaths [@Rappaport2000], ..., and ..., are often aggregated at the county level.

@@

We collected "best tracks" data on hurricane tracks for Atlantic basin storms between 1988 and 2014 from the extended best tracks database. This dataset is based on a post-storm assessment of each storm conducted by the United States National Hurricane Center (NHC) and incorporates data from a variety of sources, including satellite data and, when available, aircraft reconnaisance data [@Landsea2013]. This data gives time stamps for each observation in Coordinated Universal Time (UTC; also known as Zulu Time, sometimes indicated by "Z").

@@

We calculated the distance between each county's center and each of the 15-minute-interval estimates of a storm's location. To do this, we used the Great Circle method to calculate the distance between pairs of latitude and longitude coordinates, using the R package `sp`. For each county, we used the population mean center, based on the U.S. Census's 2010 Decennial Census, as the center location of the county. From these distance calculations, we identified the date-time and distance between the storm track and the county center at the 15-minute interval when the storm was closest to the county center. We identified the date when the storm was closest to each county and converted this date-time  to the county's local time zone using the `countytimezones` R package [citation]. This "closest date" was used to pair the distance estimates with rainfall estimates and can also be used to merge exposure datasets with daily aggregated outcomes like daily county-level mortality and morbidity counts.

@@

NLDAS does not provide information on precipitation over the ocean, preventing its use in assessing the development of the storm prior to landfall [@Villarini2011]. This point is moot for this analysis.

@@

#### Are exposure patterns different when using distance-, rain-, and wind-based metrics?

One interesting contrast is between storms that were noteworthy for direct wind-related deaths (Anderew in 1992, Hugo in 1989) versus storms noteworthy for direct freshwater flooding-related deaths (Floyd in 1999, Alberto in 1994, Charley in 1998) for the 1970--1999 period [@Rappaport2000]. Here are two examples, one of each of these two types:

```{r fig.width = 6, fig.height = 8, echo = FALSE}
storm <- "Hugo-1989"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "distance")) + 
        ggtitle(paste(storm, "distance"))
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "rainfall")) + 
        ggtitle(paste(storm, "rainfall"))

grid.arrange(a, b, ncol = 1)
```

```{r fig.width = 6, fig.height = 8, echo = FALSE}
storm <- "Floyd-1999"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "distance")) + 
        ggtitle(paste(storm, "distance"))
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "rainfall")) + 
        ggtitle(paste(storm, "rainfall"))

grid.arrange(a, b, ncol = 1)
```

#### Which storms exposed the most counties when using distance-, rain-, and wind-based metrics? What were the top county-storm exposures for rain- and wind-based metrics?

#### Are storms with heavy rain exposure likely to occur early in the hurricane season?

@@

There are a number of limitations to using distance to assess exposure to tropical storms. Tropical storms vary in size and intensity, and a measure of distance from the storm track will not incorporate these differences and so could misclassify exposure both in terms of generating false positives (counties close to the storm track of very mild or small-radius storms) and false negatives (e.g., during very large or intense storms). While a number of storm characteristics are strongly associated with distance from the storm's center (e.g., wind and, at the coast, storm surge and waves; @Rappaport2000), other characteristics like rain can occur at dangerous levels well away from the storm's central track [@Rappaport2000]. Although storm surge was a major cause of hurricane-related deaths in the past, more recently freshwater flooding has become a much more common cause of deaths directly resulting from tropical storms in the US [@Rappaport2000].

Further, the forces of a storm tend to be distributed around the center in a non-symmetrical way [citation], while distance-based exposure metrics that use buffers tend to use an equal buffer distance on each side of the storm track (e.g., [citations]).

@@

Rain from tropical storms can bring important risks to human health and property, even in counties well inland from the coast. For example, most of the direct hurricane-related deaths in the US from 1970 to 1999 were caused by freshwater flooding and were in inland, rather than coastal, counties [@Rappaport2000]. Rains from tropical storms can flood roads [@Rappaport2000].

@@

Exposure classification based on rainfall has some advantages. For example, it allows the identification of exposed counties that are inland, rather than coastal, but that were affected by heavy rainfall during the storm. Often, storm-related deaths are associated with inland flooding. In the United States, over half of hurricane-related direct deaths from 1970 to 1999 from Atlantic basin storms were a result of freshwater flooding; as a result, most of these deaths were in inland, rather than coastal, counties [@Rappaport2000]. In another analysis, researchers found that 79\% of freshwater-drowning fatalities occured in inland counties,[@Czajkowski2011] further stressing the importance of rainfall in hurricane risk analysis. Storms can produce a lot of rain especially in certain topographies, like near mountains, so counties that are well inland can sometimes experience more extreme rain that other counties at similar distances from the storm's track.

@@

Several storm characteristics are associated with the likelihood of heavy rain and flooding from the storm, including the storm's forward speed and size [@Rappaport2000]. There is some evidence that the worst tropical storms, in terms of heavy rainfall and freshwater flood-related deaths in the US, tend to occur early in the hurricane season, when it is more common to get the meteorological conditions (weak steering currents) that lead to slow-moving storms [@Rappaport2000].

@@

A tropical storm's high winds can bring a number of dangers, including health risks and property damage caused by structural damage of houses and other buildings, falling trees, and wind-borne debris [@Rappaport2000]. While severe winds pose an important threat, risks of a tropical storm can exist without severe wind; for example, one study found that most of the direct hurricane-related deaths in the US between 1970 and 1999 occurred in cases when wind was below hurricane strength (i.e., < 64 knots) [@Rappaport2000].

### 2016-06-18 writing/DraftExposurePaper/DraftExposurePaper.tex
Work on paper draft

Given its high resolution and consistancy with rain gage values,
(Villarini et al. 2011) recommends the use of Stage IV precipitation
data for the analysis of rainfall distribution in landfalling tropical
cyclones. However, the Stage IV dataset only provides data since 2001
(Lin and Mitchell 2005) and is therefore inadequate for use in this
analysis.

@@

@@

@@

In comparison to the storm track, the areas of extreme rain and extreme
wind can vary. For example, storms undergoing an extratropical
transition can bring heavy rains to the left of the storm's center
track, while extreme winds are more common to the right of the track
(Halverson 2015, Grabich et al. (2016) GBA-- I think it would be better
to just use meteorological papers as references for this point. The
Halverson paper is great, but the Grabich paper I think might just be a
secondary, rather than primary, reference for this point, so I think we
could probably find a more direct reference). Wind speeds are typically
more severe to the right of the storm's track because here the
counter-clockwise cyclonic {[}right word?{]} winds are moving in the
same direction as the forward motion of the storm, creating an additive
effect for total windspeeds {[}citation{]}. In contrast, rains can be
heaviest to the left of the storm's track, especially in cases when the
storm interacts with a downstream ridge (Atallah, Bosart, and Aiyyer
2007).

>>>>

Given its better consistancy with rain gage values compared to NLDAS
precipitation data {[}GBA-- you had something in here about resolution,
as well. Does it have a better resolution than the NLDAS data? I think
we probably want to focus on comparing this source with NLDAS, and we
don't need to put much about advantages that it has that are very
similar to NLDAS{]}, (Villarini et al. 2011) recommends Stage IV
precipitation data for the analysis of rainfall distribution in
landfalling tropical cyclones. However, the Stage IV dataset has only
been archived since 2002 (Lin and Mitchell 2005) and cannot be used to
analyze exposure to tropical storms for studies that include storms
before 2002.

@@

One interesting contrast is between storms that were noteworthy for
direct wind-related deaths (Anderew in 1992, Hugo in 1989) versus storms
noteworthy for direct freshwater flooding-related deaths (Floyd in 1999,
Alberto in 1994, Charley in 1998) for the 1970--1999 period (Rappaport
2000). Here are two examples, one of each of these two types:

@@

\includegraphics{DraftExposurePaper_files/figure-latex/unnamed-chunk-3-1.pdf}

@@

In comparison to the storm track, the areas of extreme rain and extreme
wind can vary. For example, storms undergoing an extratropical
transition can bring heavy rains to the left of the storm's center track
(Halverson 2015), while extreme winds are more common to the right of
the track (Halverson 2015, Grabich et al. (2016) GBA-- I think it would
be better to just use meteorological papers as references for this
point. The Halverson paper is great, but the Grabich paper I think might
just be a secondary, rather than primary, reference for this point, so I
think we could probably find a more direct reference). Wind speeds are
typically more severe to the right of the storm's track because here the
counter-clockwise cyclonic {[}right word?{]} winds are moving in the
same direction as the forward motion of the storm, creating an additive
effect for total windspeeds {[}citation{]}. In contrast, rains can be
heaviest to the left of the storm's track, especially in cases when the
storm interacts with a downstream ridge (Atallah, Bosart, and Aiyyer
2007).
